---
title: "BIOST 561 Homework 1"
author: "Hantong Hu"
date: "04/16/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
### Setting up the packages, options we'll need:
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
### -----------------------------------------------------------
### Reading in the data. 
# library(tidyverse)
# mri_data <- read_csv("C:/Users/second/Desktop/BIOST 515/HW/Datasets/mri-2021.csv")
```

# Responses 

## Problem 2

### 2.1  

my_list[["a1"]]: returns the content of a1, which is b1 and b2 under a1.

my_list[["a1"]][["b2"]]: returns the content of b2, which is c1 and c2 under b2.

my_list[["a1"]][["b2"]][["c1"]]: returns the content of c1 under a1 and b2, which is "a1b2c1".

my_list[[c(1,2)]]: returns the content of b2, which is c1 and c2 under b2.

my_list[[c(1,2,1)]]: returns the content of c1 under a1 and b2, which is "a1b2c1".

my_list[[c("a1", "b2")]]: returns the content of b2, which is c1 and c2 under b2.

my_list[[c("a1", "b2", "c1")]]: returns the content of c1 under a1 and b2, which is "a1b2c1".

my_list[["a1"]][["b2"]], my_list[[c(1,2)]], my_list[[c("a1", "b2")]] are equivalent.

my_list[["a1"]][["b2"]][["c1"]], my_list[[c(1,2,1)]], my_list[[c("a1", "b2", "c1")]] are equivalent.

```{r Q2.1, message=F, results='hide'}
### -----------------------------------------------------------
### Q2.1
my_list <- list(
  a1 = list(b1 = list(c1="a1b1c1", c2="a1b1c2"),
            b2 = list(c1="a1b2c1", c2="a1b2c2")),
  a2 = 1
)

my_list[["a1"]]
my_list[["a1"]][["b2"]]
my_list[["a1"]][["b2"]][["c1"]]
my_list[[c(1,2)]]
my_list[[c(1,2,1)]]
my_list[[c("a1", "b2")]]
my_list[[c("a1", "b2", "c1")]]
```

### 2.2

my_list[c("a1", "b2", "c1")] returns the 3 elements (a1, b2, c1) of my_list, but since my_list only has elements a1 and a2, this line can only return a1 and the other 2 elements are returned as NA.

my_list[[c("a1", "b2", "c1")]] returns the content of c1 under a1 and b2, so it will return "a1b2c1".

my_list[c(1,2)] returns the first and second elements (a1, a2) of my_list.

my_list[[c(1,2)]] returns the content of b2 under a1, so it will return c1 and c2 under b2 and a1.

```{r Q2.2, message=F, results='hide'}
### -----------------------------------------------------------
### Q2.2
my_list[c("a1", "b2", "c1")]
my_list[[c("a1", "b2", "c1")]]

my_list[c(1,2)]
my_list[[c(1,2)]]
```

## Problem 3

### 3.1  

```{r Q3.1, message=F}
### -----------------------------------------------------------
### Q3.1
visits <- data.frame(patient_id = c(1:5,5),
                     treatment_id = c("a","b","a","c","a","c"),
                     n_visits = c(2,1,2,1,3,1))
patients <- data.frame(patient_id = 1:6, name = LETTERS[1:6], age = rpois(6,50))
treatments <- data.frame(patient_id = c("a","b","c"), name = tail(letters,3),
                         cost = c(3,1000,100))

X_ids <- data.frame(visits[c(rep(c(1:6), times = c(visits$n_visits))),c(1,2)])
X_ids
```

### 3.2 

```{r Q3.2, message=F}
### -----------------------------------------------------------
### Q3.2
X <- data.frame(patient_name = patients[match(X_ids$patient_id, patients$patient_id),"name"],
                treatment_name = treatments[match(X_ids$treatment_id, treatments$patient_id),"name"])
X$age <- patients[match(X$patient_name, patients$name),"age"]

patients
X
```

### 3.3 

```{r Q3.3, message=F}
### -----------------------------------------------------------
### Q3.3
X_order <- X[order(X$age),]
X_order
```

### 3.4

```{r Q3.4, message=F}
### -----------------------------------------------------------
### Q3.4
visit_for_trt <- function(trt){
  return(X[which(X$treatment_name==trt),])
}

visit_for_trt("x")
```

## Problem 4

### 4.1  

state.x77 is a matrix with double type data.

```{r Q4.1, message=F, warning=F}
### -----------------------------------------------------------
### Q4.1
library(MASS)
data(state) # Load dataset

class(state.x77)
typeof(state.x77)
```

### 4.2


```{r Q4.2, message=F}
### -----------------------------------------------------------
### Q4.2
mean <- apply(state.x77, 2, mean)
std <- apply(state.x77, 2, sd)

mat4.2 <- rbind(mean, std)
mat4.2
```

### 4.3

lapply returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X.

sapply is a user-friendly version and wrapper of lapply by default returning a vector, matrix or, if simplify = "array", an array if appropriate.

```{r Q4.3, message=F}
### -----------------------------------------------------------
### Q4.3
state.x77.df <- as.data.frame(state.x77)

mean_df <- sapply(state.x77.df, mean)
std_df <- sapply(state.x77.df, sd)
mat4.3 <- rbind(mean_df, std_df)
mat4.3
```

### 4.4

From the system process time, adding a column to a matrix is more efficient than to a data frame. It might be due to that data frame needs to check the length of vector before adding the column but matrix does not need to.

```{r Q4.4, message=F}
### -----------------------------------------------------------
### Q4.2
start.df <- Sys.time()
state.x77.df$Density <- state.x77.df$Population / state.x77.df$Area
end.df <- Sys.time()
df.diff <- end.df - start.df

start.mat <- Sys.time()
state.x77.dens <- cbind(state.x77, state.x77[,"Population"] / state.x77[,"Area"])
end.mat <- Sys.time()
mat.diff <- end.mat - start.mat

paste("Data Frame process time is", df.diff, "secs")
paste("Matrix process time is", mat.diff, "secs")
```

### 4.5


```{r Q4.5, message=F}
### -----------------------------------------------------------
### Q4.5
state.x77.df$region <- state.region

max.dens <- tapply(state.x77.df$Density, state.x77.df$region, FUN = max)
min.dens <- tapply(state.x77.df$Density, state.x77.df$region, FUN = min)
mean.dens <- tapply(state.x77.df$Density, state.x77.df$region, FUN = mean)

rbind(max.dens, min.dens, mean.dens)
```

## Problem 5

```{r Q5, message=F}
### -----------------------------------------------------------
### Q5
`modif<-` <- function(y,value) {
  y[which(y %in% NA)] <- value
  y
}

x <- c(1:5,NA,7,8,NA,10)
modif(x) <- 2
x
```

## Problem 6

A variable a is created in the global environment and the value 10 is assigned to it.

The variable pi is searched in the global environment, but then a variable x is created in the global environment and the value pi is assigned to it.

A variable z is created in the global environment and the value 10 is assigned to it.

Variables x and n are created and assigned value (of 2 and 2 in this case) in the execution environment. 
A new value from calculations of x=2 and n=2 is then assigned to x (still in the execution environment). 
A variable z is created in the execution environment and the value of x is assigned to it.

A variable w is created in the execution environment and the value of z*a (z is in the execution environment and a is in the global environment) is assigned to it.

```{r Q6, message=F}
### -----------------------------------------------------------
### Q6
rm(list = ls())

a <- 10
x <- pi
z <- 20

fn <- function(x,n){
  x <- x^n
  z <- x
  w <- z*a
  return(w)
}


fn(2,2)
```

\pagebreak

## Code Appendix

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
